---
title: "Harkins_Final_Project"
author: "Sarah Harkins"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library(deSolve)
library(ggplot2)
library(bbmle)
library(MASS)
```

```{r}
# Load in the pathogen data
black_data = read.csv("/Users/harkins/Desktop/560/black_0.05_mg_per_liter.csv")
green_data = read.csv("/Users/harkins/Desktop/560/green_control.csv")
red_data = read.csv("/Users/harkins/Desktop/560/red_0.01_mg_per_liter.csv")

head(black_data)
head(red_data)
head(green_data)
```
```{r}
#combining data frames to one mega-data frame
observed_df = data.frame(black_data)

observed_df$red_data_day <- red_data$day
observed_df$red_data_parasite <- red_data$parasite

observed_df$green_data_day <- green_data$day
observed_df$green_data_parasite <- green_data$parasite

#observed_df
```

```{r}
install.packages("ggplot2") # If you don't already have ggplot
library(ggplot2)

ggplot(data=observed_df, aes(x=day, y=parasite, color="0.05 mg l^{-1}")) + geom_line() + 
            geom_point() + 
  geom_point(aes(x=red_data_day, y=red_data_parasite, color="0.01 mg l^{-1}")) +
            geom_line(aes(x=red_data_day, y=red_data_parasite, color="0.01 mg l^{-1}")) + 
  geom_point(aes(x=green_data_day, y=green_data_parasite, color="Control")) +
            geom_line(aes(x=green_data_day, y=green_data_parasite, color="Control"))

```
```{r}
#considering the exponentiated observed (the mean parasite intensity)
exp_observed_df = data.frame(matrix(ncol = 0, nrow = 14))
exp_observed_df$day_black = observed_df$day
exp_observed_df$exp_black = exp(observed_df$parasite)

exp_observed_df$day_red = observed_df$red_data_day
exp_observed_df$exp_red = exp(observed_df$red_data_parasite)

exp_observed_df$day_green = observed_df$green_data_day
exp_observed_df$exp_green = exp(observed_df$green_data_parasite)

exp_observed_df

exp_control_rounded = round(exp_observed_df$exp_green,0)
exp_control_rounded
```

```{r}
ggplot(data=exp_observed_df, aes(x=day_black, y=exp_black, color="0.05 mg l^{-1}")) + geom_line() + 
            geom_point() + 
  geom_point(aes(x=day_red, y=exp_red, color="0.01 mg l^{-1}")) +
            geom_line(aes(x=day_red, y=exp_red, color="0.01 mg l^{-1}")) + 
  geom_point(aes(x=day_green, y=exp_green, color="Control")) +
            geom_line(aes(x=day_green, y=exp_green, color="Control"))
```




```{r}
pi_model_HT1 = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  beta = params["beta"]
  d = params["d"] #death term of immune particles
  p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  f = params["f"] #rate at which the immune particles are produce due to detected parasite 
  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: the parasite measure in mass 
  
  #considering a Holling Type I functional response
  dP = beta*P - p*I*P
  dI = B - d*P +f*P*I   

  #try to show delayed pred/prey wave between parasite and immune system 
  # capture one "wave" over the 30 day experiment 
  
  return(list(c(dP, dI)))
}
```

```{r}
pi_model_HT2 = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  beta = params["beta"]
  d = params["d"] #death term of immune particles [immune particles per parasite per time]
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 
  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: the parasite measure in mass 
  
    #considering a Holling Type II function response 
  dP = beta*P - I*P*c/(a+P)
  dI = B - d*P + P*I*b/(a+P) 
 
  return(list(c(dP, dI)))
}
```

```{r}
pi_model_HT3 = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 

  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 
  
  #considering a Holling Type III function response 
  dP = beta*P - I*P^2*c/(a^2+P^2)
  dI = B - d*P + I*P^2*b/(a^2+P^2)

  return(list(c(dP, dI)))
}
```

```{r}
#times = seq(0, 30, len = 100) 

# Initial conditions
#init_vals = c(P= observed_df$green_data_parasite[1], I = 0.2)

#guessed parameters
#params = c(beta=0.11, d = -0.2, p = 0.3 , B = -0.005, a = 2.6, c = 0.7) #Gamma = 0.05, c = 0.001, i = 0.005, a = 5, m=0.15

#beta: ?Proportionality constant
#Gamma: rate at which guppies acquire parasite 
#c: max per capita consumption rate of guppies by parasites (should be small number)
#i:immunity of parasite to guppy defense (immune system)
#a:"shape parameter" 1/2 saturation rate in absence of inhibitory effect, i
#m: mortality rate of parasite
 

#sim_0.05 = ode(init_vals, times, pi_model, params) 
#sim_0.05_df = data.frame(sim_0.05)
#sim_0.05_df = log(sim_0.05_df)
#sim_0.05_df

```


```{r}
ggplot() + geom_line(aes(x=sim_0.05_df$time, y=sim_0.05_df$P, color='Sim. Control Parasites'))+
  geom_line(aes(x=sim_0.05_df$time, y=sim_0.05_df$I, color='Sim. Control Immune Particles'))+ 
    geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Log Parasite Intensity") + xlab("Time")
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT1 = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT1, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT2 = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT2, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT3 = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT3, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```


Finding optimal parameters for each system of ODEs then plotting the simulation with those parameters against the observed data from the paper. 

```{r}
#using optim to find appropriate parameters
params_HT1 = c(beta=0.11, d = -0.2, p = 0.3 , B = -0.005, f=0.1)

#FOR HOLLINGS TYPE I
fit0_HT1 = optim(params_HT1, rss_pi_HT1, data=observed_df$green_data)
#fit0_HT1$par
#fit0$convergence
fit1_HT1 = optim(fit0_HT1$par, rss_pi_HT1, data=observed_df$green_data)
#fit1_HT1$par
fit3_HT1 = optim(fit1_HT1$par, rss_pi_HT1, data=observed_df$green_data)
fit3_HT1$par
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT1 = ode(init_vals, times, pi_model_HT1, fit3_HT1$par)

pi_results_min_df_HT1 = data.frame(pi_results_min_HT1)
pi_results_min_df_HT1

ggplot() + geom_line(aes(x=pi_results_min_df_HT1$time, y=pi_results_min_df_HT1$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```


```{r}
params_HT2 = c(beta=0.11, d = -0.2 , B = -0.005, a = 2.6, c = 0.7, b=0.1)


#FOR HOLLINGS TYPE II
fit0_HT2 = optim(params_HT2, rss_pi_HT2, data=observed_df$green_data)
#fit0_HT2$par
#fit0$convergence
fit1_HT2 = optim(fit0_HT2$par, rss_pi_HT2, data=observed_df$green_data)
#fit1_HT2$par
fit3_HT2 = optim(fit1_HT2$par, rss_pi_HT2, data=observed_df$green_data)
fit3_HT2$par
```

```{r}
#RSS of the Holling type II model 
rss_pi_HT2(fit3_HT2$par,observed_df$green_data )
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT2 = ode(init_vals, times, pi_model_HT2, fit3_HT2$par)

pi_results_min_df_HT2 = data.frame(pi_results_min_HT2)
#pi_results_min_df_HT2

ggplot() + geom_line(aes(x=pi_results_min_df_HT2$time, y=pi_results_min_df_HT2$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```

```{r}
params_HT3 = c(beta=0.11, d = -0.2 , B = -0.005, a = 2.6, c = 0.7, b=0.1)

#FOR HOLLINGS TYPE III
fit0_HT3  = optim(params_HT3, rss_pi_HT3, data=observed_df$green_data)
#fit0_HT3$par
#fit0$convergence
fit1_HT3 = optim(fit0_HT3$par, rss_pi_HT3, data=observed_df$green_data)
#fit1_HT3$par
fit3_HT3 = optim(fit1_HT3$par, rss_pi_HT3, data=observed_df$green_data)
fit3_HT3$par
```

```{r}
#RSS of the Holling type III model 
rss_pi_HT3(fit3_HT3$par, observed_df$green_data)
```


```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT3 = ode(init_vals, times, pi_model_HT3, fit3_HT3$par)

pi_results_min_df_HT3 = data.frame(pi_results_min_HT3)
#pi_results_min_df_HT3

ggplot() + geom_line(aes(x=pi_results_min_df_HT3$time, y=pi_results_min_df_HT3$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```

Now, we will calculate AIC to determine which of these three models is best suited. We will consider Poisson measurement error. 

```{r}
HT1_nll = function(beta, d, p, B, f){
  
  # Setting initial conditions and parameters just like before
  times = observed_df$green_data_day
  params = c(beta=beta, d=d, p=p, B=B, f=f)
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT1, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  ode_res
  nll = -1*sum(dpois(observed_df$green_data_parasite, ode_res$P[2:15], log=TRUE)) #observed_df$green_data_parasite
  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit3_HT1$par)
start_params

HT1_nll(start_params[1], 
         start_params[2], 
         start_params[3],
         start_params[4],
         start_params[5])
```

```{r}
library(bbmle)

# Get the mle estimates for my parameters
fit_all_HT1 = mle2(HT1_nll, 
               start=list(beta = start_params[1], 
                          d = start_params[2], 
                          p = start_params[3],
                          B = start_params[4],
                          f = start_params[5]),
               method="L-BFGS-B",
               lower=c(0, 0, 0,0,0),
               upper=c(Inf, Inf, Inf, Inf, Inf))


AIC(fit_all_HT1)


```






To prevent unbounded growth of the parasite, seen in the simulations of the Hollings Type I and II models, we will impose a carrying capacity via logistic growth in the growth term of the parasites equation. 

```{r}
pi_model_HT1_car_cap = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  #beta = params["beta"]
  d = params["d"] #death term of immune particles
  p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  K = params["K"] #carrying capacity of parasites in the tank 
  r = params["r"]
  f = params["f"]
  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 
  
  #considering a Holling Type I functional response
  dP = r*P*(1-P/K) - p*I*P
  dI = B + f*P*I - d*P   

  return(list(c(dP, dI)))
}
```


```{r}
pi_model_HT2_car_cap = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  #beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  K = params["K"] #carrying capacity of parasites in the tank 
  r = params["r"] #intrinsic growth rate of parasite 
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 

  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 
  
  #considering a Holling Type II function response 
  dP = r*P*(1-P/K) - I*P*c/(a+P)
  dI = B - d*P + P*I*b/(a+P)

  return(list(c(dP, dI)))
}
```

```{r}
pi_model_HT3_car_cap = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  #beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  K = params["K"] #carrying capacity of parasites in the tank 
  r = params["r"] #intrinsic growth rate of parasite 
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 

  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 

  #considering a Holling Type III function response 
  dP = r*P*(1-P/K) - I*P^2*c/(a^2+P^2)
  dI = B - d*P + I*P^2*b/(a^2+P^2)
  


  return(list(c(dP, dI)))
}
```


```{r}
rss_pi_HT1_car_cap = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT1_car_cap, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT2_car_cap = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT2_car_cap, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT3_car_cap = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT3_car_cap, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```


```{r}
#using optim to find appropriate parameters
params_HT1_car_cap = c(d = 0.2, p = 0.3 , B = 0.005, K=10, r=0.1, f=0.1)

#FOR HOLLINGS TYPE I
fit0_HT1_car_cap = optim(params_HT1_car_cap, rss_pi_HT1_car_cap, data=observed_df$green_data)
fit0_HT1_car_cap$par
#fit0$convergence
fit1_HT1_car_cap = optim(fit0_HT1$par, rss_pi_HT1_car_cap, data=observed_df$green_data)
fit1_HT1_car_cap$par
fit3_HT1_car_cap = optim(fit1_HT1_car_cap$par, rss_pi_HT1_car_cap, data=observed_df$green_data)
fit3_HT1_car_cap$par
```
```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT1_car_cap = ode(init_vals, times, pi_model_HT1_car_cap, fit3_HT1$par)

pi_results_min_df_HT1_car_cap = data.frame(pi_results_min_HT1_car_cap)
pi_results_min_df_HT1_car_cap
ggplot() + geom_line(aes(x=pi_results_min_df_HT1_car_cap$time, y=pi_results_min_df_HT1_car_cap$P, color='Optimized Parameters'))


ggplot() + geom_line(aes(x=pi_results_min_df_HT1_car_cap$time, y=pi_results_min_df_HT1_car_cap$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```


```{r}
params_HT2_car_cap = c(beta=0.11, d = -0.2, p = 0.3 , B = -0.005, a = 2.6, c = 0.7, K=1000, r = 0.01, b=0.1)


#FOR HOLLINGS TYPE II
fit0_HT2_car_cap = optim(params_HT2_car_cap, rss_pi_HT2_car_cap, data=observed_df$green_data)
fit0_HT2_car_cap$par
#fit0$convergence
fit1_HT2_car_cap = optim(fit0_HT2_car_cap$par, rss_pi_HT2_car_cap, data=observed_df$green_data)
fit1_HT2_car_cap$par
fit3_HT2_car_cap = optim(fit1_HT2_car_cap$par, rss_pi_HT2_car_cap, data=observed_df$green_data)
fit3_HT2_car_cap$par
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT2_car_cap = ode(init_vals, times, pi_model_HT2_car_cap, fit3_HT2_car_cap$par)

pi_results_min_df_HT2_car_cap = data.frame(pi_results_min_HT2_car_cap)
#pi_results_min_df_HT2

ggplot() + geom_line(aes(x=pi_results_min_df_HT2_car_cap$time, y=pi_results_min_df_HT2_car_cap$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```

```{r}
params_HT3_car_cap = c(beta=0.11, d = -0.2, p = 0.3 , B = -0.005, a = 2.6, c = 0.7, K=1000, r = 0.01, b=0.1)

#FOR HOLLINGS TYPE III
fit0_HT3_car_cap = optim(params_HT3_car_cap, rss_pi_HT3_car_cap, data=observed_df$green_data)
fit0_HT3_car_cap$par
#fit0$convergence
fit1_HT3_car_cap  = optim(fit0_HT3_car_cap$par, rss_pi_HT3_car_cap, data=observed_df$green_data)
fit1_HT3_car_cap$par
fit3_HT3_car_cap = optim(fit1_HT3_car_cap$par, rss_pi_HT3_car_cap, data=observed_df$green_data)
fit3_HT3_car_cap$par
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT3 = ode(init_vals, times, pi_model_HT3_car_cap, fit3_HT3$par)

pi_results_min_df_HT3 = data.frame(pi_results_min_HT3)
#pi_results_min_df_HT3

ggplot() + geom_line(aes(x=pi_results_min_df_HT3$time, y=pi_results_min_df_HT3$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```








______________________________________________________________________________________________


```{r}
#considering the optimized parameters for the system of ODEs using the exponentiated observed data set
fit2 = optim(params, rss_pi, data=exp_observed_df$green_data)
fit3 = optim(fit2$par, rss_pi, data=exp_observed_df$green_data)
fit3$par
```

```{r}
control_expond = data.frame(matrix(ncol = 0, nrow = 14))
control_expond$control_day = observed_df$green_data_day
control_expond$control_expond_para = exp(observed_df$green_data_parasite)
```



```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals_2 = c(P= control_expond$control_expond[1], I = 0.1)

#visual guess parameters
pi_results_guess_2 = ode(init_vals_2, times, pi_model, params)
sibr_results_guess_df_2 = data.frame(pi_results_guess_2)

#parameters that minimize the RSS
pi_results_min_2 = ode(init_vals_2, times, pi_model, fit1$par)
pi_results_min_df_2 = data.frame(pi_results_min_2)

ggplot() + geom_line(aes(x=sibr_results_guess_df_2$time, y=sibr_results_guess_df_2$P, color='Guess Parameters'))+
          geom_line(aes(x=pi_results_min_df_2$time, y=pi_results_min_df_2$P, color='Optimized Parameters'))+
      geom_point(aes(x=control_expond$control_day, y=control_expond$control_expond_para, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=control_expond$control_day, y=control_expond$control_expond_para, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```









The oringial plot looks like a gamma distriubiton. 

After some thinking, the gamma distribution won't be the best fit because it crosses at (0,0) and the data does not. I am considering the skewed right normal distribution. I do not believe this has a closed form equaiton so it may be difficult to fit parameters to it for that reason. 
```{r}

x_samp = seq(0, 30, len = 100)
gamma_sample = dgamma(x_samp, shape=5, rate = 0.5,  log = FALSE) #scale = 1/rate,

ggplot() + geom_line(aes(x=x_samp, y=gamma_sample, color='0.05'))+
  ylab("Parasite Intensity") + xlab("Time")
```

