---
title: 'Lecture code: Multivariate Regression'
author: "Mark Wilber"
date: "`r Sys.Date()`"
output: html_document
---

```{r}
library(ggplot2)
library(data.table) # I don't use tidyverse, but do your own thing if you want
library(dagitty)
library(GGally)
```

```{r}
library(GGally)
svl_data = read.csv("tadpole_svl.csv")
head(svl_data)
svl_data$svl

ggpairs(svl_data)
ggsave("cor_plot.pdf", width=10, height=4)
```

```{r}
# Standardize algae
svl_data$algae_z = with(svl_data, (algae - mean(algae)) / sd(algae))
svl_data$algae_z
```

What is standardization doing?

```{r}
library(patchwork)

n = 10000
x = rnorm(n, mean=20, sd=2)
x_z = (x - mean(x)) / sd(x)
tdat = data.frame(vals=c(x, x_z), 
				 type=rep(c("Original", "Standardized"), c(n, n)))
ggplot(data=tdat) + geom_histogram(aes(x=vals)) + facet_wrap(~type, scales='free') + theme_classic() + xlab("x values")

# ggsave("compare_hists.pdf", width=8, height=4)


```

```{r}
library(data.table)

# Standardize
svl_data$algae_z = with(svl_data, (algae - mean(algae)) / sd(algae))
svl_data$svl_z = with(svl_data, (svl - mean(svl)) / sd(svl))

# Prior simulation
n = 50
beta0 = rnorm(n, 0, 2)
beta1 = rnorm(n, 0, 2)
vals = seq(min(svl_data$algae_z), max(svl_data$algae_z), len=100)

prior_lines = sapply(1:n, function(i) beta0[i] + beta1[i]*vals)
prior_dt = data.table(prior_lines)
prior_dt$algae_z = vals
prior_melt = melt(prior_dt, id.vars="algae_z")
ggplot(prior_melt) + geom_line(aes(x=algae_z, y=value, group=variable)) + theme_classic() + xlab("algae_z") + ylab("svl_z")

#ggsave("prior_simulations.pdf", width=4, height=3)

```

Fit the model with standardized variables

```{r}
library(rethinking)
svl_data$algae_z = with(svl_data, (algae - mean(algae)) / sd(algae))
svl_data$svl_z = with(svl_data, (svl - mean(svl)) / sd(svl))

mod1 = quap(
		alist(
			  svl_z ~ dnorm(mu, sigma),
			  mu <- beta0 + beta1*algae_z,
			  beta0 ~ dnorm(0, 2),
			  beta1 ~ dnorm(0, 2),
			  sigma ~ dexp(1)
				), data=svl_data)
precis(mod1, prob=0.95)

# Plot the posteriors
post = extract.samples(mod1, n=10000)
beta1 = post[, 2] # effect size 
sd_svl = sd(svl_data$svl); sd_alg = sd(svl_data$algae)
beta1_unstd = beta1*(sd_svl / sd_alg) # posterior for unstandardized (untransformed) scale 
precis(beta1_unstd, prob=0.95)

pdf("density_plot.pdf", width=5, height=3)
dens(post[, -3])
dev.off()

# Check that fitting the de-meaned algae gives the same result
mod2 = quap(
		alist(
			  svl ~ dnorm(mu, sigma),
			  mu <- beta0 + beta1*(algae - mean(algae)),
			  beta0 ~ dnorm(67, 4),
			  beta1 ~ dnorm(0, 20),
			  sigma ~ dunif(0, 100)
				), data=svl_data)
mod2

# Also check with a frequentist approach
summary(lm(svl ~ algae, data=svl_data))

```

Check the single variable model

```{r}
library(patchwork)

# Check linearity
algae_pred = seq(min(svl_data$algae_z),
				 max(svl_data$algae_z), 
				 len=100)
# Simulate predictions
pred_dt = data.frame(algae_z=algae_pred)
pred = link(mod1, data=pred_dt)
pred_dt$med = apply(pred, 2, median)
pred_dt$lower =  apply(pred, 2, quantile, 0.025)
pred_dt$upper =  apply(pred, 2, quantile, 0.975)

ggplot(svl_data) + geom_point(aes(x=algae_z, y=svl_z)) + 
				   geom_line(data=pred_dt, aes(x=algae_z, y=med), color='blue') +
				   geom_ribbon(data=pred_dt, aes(x=algae_z, ymin=lower, ymax=upper), fill='blue', alpha=0.1) +
				   theme_classic()

# ggsave("mod1_predictions.pdf", width=4, height=3)
```

```{r}
# Median residuals
pred2 = link(mod1, n = 1,000) #predicted posterior distribution for each point in data set 
                              # n=1,000 is the default number of samples per point 

hist(pred2[,1]) # predicted of svl_z posterior distribution of mean for first data point

mean_pred = colMeans(pred2) # mean across all of the data 
resid = svl_data$svl_z - mean_pred # compute residual (observed - predicted svl)

# QQplot to check normality
df = data.frame(resid=resid, x=svl_data$algae_z)
p = ggplot(df, aes(sample = resid))
pq = p + stat_qq() + stat_qq_line() + theme_classic()

# Residual plot to check equal variance
pr = ggplot(df, aes(x=x, y=resid)) + geom_point() + geom_hline(aes(yintercept=0), linetype="dashed") + xlab("algae_z") + ylab("residual") + theme_classic()
pq + pr
# ggsave("resid_plots.pdf", width=7, height=3)


```

```{r}

# Calculate the Bayesian R2 with uncertainty
pred2 = link(mod1, n=10000)
all_resid = t(svl_data$svl_z - t(pred2)) # calculate residual for all points in posterior distribution
resid_var = apply(all_resid, 1, var)
mean_var =  apply(pred2, 1, var)
R2 = mean_var / (mean_var + resid_var) # this is a distribution! 

median(R2) # reportable conclusion along with the 95% confidence/credible interval

ggplot() + geom_histogram(aes(x=R2)) + 
		   geom_vline(aes(xintercept=median(R2)), linetype="dashed", linewidth=1) +
		   theme_classic()

# ggsave("bayesian_r2.pdf", width=4, height=3)


```

```{r}

# Classic biostats

full_mod = lm(svl ~ density*area + algae*nutr, data=svl_data)
best = step(full_mod, direction="both") # drops anf includes variables to find the one with the lowest AIC, thus producing the best model 
summary(best, direction="both")

```


Making DAGs in R

```{r}
library(dagitty)
# Draw example DAG
dag_ex <- dagitty("dag{
Nutr -> SVL
Nutr -> Algae
Algae -> SVL
SVL [outcome]
}")

coordinates(dag_ex) <- list(x=c(Nutr=1, Algae=3, SVL=2),
						  y=c(Nutr=0, Algae=0, SVL=3))

pdf("dag_example1.pdf", width=4, height=3)
rethinking::drawdag(dag_ex)
dev.off()

dag_ex2 <- dagitty("dag{
Nutr -> SVL
Nutr -> Algae
Algae -> SVL
SVL [outcome]
}")

coordinates(dag_ex2) <- list(x=c(Nutr=1, Algae=3, SVL=2),
						  y=c(Nutr=0, Algae=0, SVL=3))

pdf("dag_example2.pdf", width=4, height=3)
rethinking::drawdag(dag_ex2)
dev.off()

dag_ex3 <- dagitty("dag{
Nutr -> SVL
Algae -> SVL
SVL [outcome]
}")

coordinates(dag_ex3) <- list(x=c(Nutr=1, Algae=3, SVL=2),
						  y=c(Nutr=0, Algae=0, SVL=3))

pdf("dag_example3.pdf", width=4, height=3)
rethinking::drawdag(dag_ex3)
dev.off()


dag_ex4 <- dagitty("dag{
GoodLooking -> Actor
Funny -> Actor
Actor [outcome]
}")

coordinates(dag_ex4) <- list(x=c(GoodLooking=1, Funny =3, Actor=2),
						  y=c(GoodLooking=0, Funny=0, Actor=3))

pdf("dag_example4.pdf", width=4, height=3)
rethinking::drawdag(dag_ex4)
dev.off()


dag_ex5 <- dagitty("dag{
Nutr -> SVL
Algae -> SVL
SVL -> Repro
}")

coordinates(dag_ex5) <- list(x=c(Nutr=1, Algae=3, SVL=2, Repro=2),
						  y=c(Nutr=0, Algae=0, SVL=2, Repro=3))

pdf("dag_example5.pdf", width=4, height=3)
rethinking::drawdag(dag_ex5)
dev.off()


```

```{r}

taddag <- dagitty("dag{
Area -> Density -> SVL
Area -> Algae -> Density
Nutr -> Algae -> SVL
Nutr -> SVL
Area [exposure]
SVL [outcome]
}")


coordinates(taddag) <- list(x=c(Area=1, Algae=5, Density=5,
                              Nutr = 5,SVL=10),
							y=c(Area=5, Algae=5, Density=9,
								Nutr = 1, SVL=5))


#pdf("tad_diag.pdf", width=4, height=3)
rethinking::drawdag(taddag)
dev.off()

# What do you need to hold constant to get the direct effect of algae
adjustmentSets(taddag, exposure="Algae", outcome="SVL", effect="direct")

# What do you need to hold constant to get the total effect of algae?
adjustmentSets(taddag, exposure="Algae", outcome="SVL", effect="total")

```


```{r}

taddag2 <- dagitty("dag{
Area -> Algae -> SVL
Area -> SVL
Nutr -> Algae
Density -> SVL
Area [exposure]
SVL [outcome]
}")

coordinates(taddag2) <- list(x=c(Area=1, Algae=5, Density=1,
                              Nutr = 1, SVL=10),
                         y=c(Area=1, Algae=5, Density=9,
                             Nutr = 5, SVL=5))

#pdf("tad_diag2.pdf", width=4, height=3)
rethinking::drawdag(taddag2)
dev.off()
```

Build a Bayesian multiple regression

```{r}
# Quick way to standardize your dataset
svl_data_z = as.data.frame(scale(svl_data))
svl_data_z

tadmod = quap(
			alist(
				svl ~ dnorm(mu, sigma),
				mu <- beta0 + beta_algae*algae + beta_nutr*nutr + beta_density*density,
				beta0 ~ dnorm(0, 5),
				beta_algae ~ dnorm(0, 2),
				beta_nutr ~ dnorm(0, 2),
				beta_density ~ dnorm(0, 2),
				sigma ~ dexp(1)
				), data=svl_data_z)
precis(tadmod, prob=0.95)

# Extract effects on the svl scale
post = extract.samples(tadmod, n=1000)
beta_nat = post[, 2] * sd(svl_data$svl) # unstandardized in the y variable 
precis(beta_nat, prob=0.95)

#pdf("algae_effect.pdf", width=4, height=3)
hist(beta_nat, xlab="Algae effect on svl (mm) ", main="")
dev.off()


# Get the total effect of algae 
tadmod2 = quap(
      alist(
        svl ~ dnorm(mu, sigma),
        mu <- beta0 + beta_algae*algae + beta_nutr*nutr + beta_area*area,
        beta0 ~ dnorm(0, 5),
        beta_algae ~ dnorm(0, 2),
        beta_nutr ~ dnorm(0, 2),
        beta_area ~ dnorm(0, 2),
        sigma ~ dexp(1)
        ), data=svl_data_z)
precis(tadmod2, prob=0.95)

```

Test your DAGs dependencies

```{r}
library(dagitty) # Package for building DAGs

# Build your DAG
dag2 <- dagitty("dag{
Area -> Density -> SVL
Area -> Algae -> Density
Nutr -> Algae -> SVL
Nutr -> SVL
}")

# Format your DAG
coordinates(dag2) <- list(x=c(Area=1, Algae=5, Density=5,
                              Nutr = 1, SVL=10),
                         y=c(Area=5, Algae=5, Density=9,
                         	Nutr = 1, SVL=5))
# Visualize DAG
rethinking::drawdag(dag2)

# Look at implied dependencies
impliedConditionalIndependencies(dag2)

```

Check that our DAG is consistent with our data

```{r}


# Condition 1: Area and Nutr should be uncorrelated

# Area is predictor variable and Nutr is response variable 
test1 = quap(
			alist(
				area ~ dnorm(mu, sigma),
				mu <- b0 + b1*nutr,
				b0 ~ dnorm(0, 5),
				b1 ~ dnorm(0, 2),
				sigma ~ dexp(1)), data=svl_data_z)
precis(test1, prob=0.95)
# the 95% confidence interval of b1 overlaps with 0 so there is not a strong correlation between the variables 

# Condition 2: Area and SVL should be uncorrelated given
# algae, density, and nutr
test2 = quap(
			alist(
				# R1
				svl ~ dnorm(mu, sigma),
				mu <- b0 + b4*algae + 
						    b3*density +
						    b2*nutr +
						    b1*area,
				b0 ~ dnorm(0, 5),
				b1 ~ dnorm(0, 2),
				b2 ~ dnorm(0, 2),
				b3 ~ dnorm(0, 2),
				b4 ~ dnorm(0, 2),
				sigma ~ dexp(1)), data=svl_data_z)
precis(test2, prob=0.95)

# Condition 3: Density and nutrient are not related after controlling
# for algae and area
test3 = quap(
			alist(
				density ~ dnorm(mu, sigma),
				mu <- b0 + b3*algae + 
						    b2*area +
						    b1*nutr,
				b0 ~ dnorm(0, 5),
				b1 ~ dnorm(0, 2),
				b2 ~ dnorm(0, 2),
				b3 ~ dnorm(0, 2),
				sigma ~ dexp(1)), data=svl_data_z)
precis(test3, prob=0.95)

# Plot the coefficient estimates
all_coefs = list()
mods = list(test1=test1, test2=test2, test3=test3)
for(i in 1:length(mods)){
	mod = mods[[i]]
	tpost = extract.samples(mod)$b1
	ests = quantile(tpost, c(0.025, 0.5, 0.975))
	tdt = data.frame(model_nm=names(mods)[i], 
					 med=ests[2], 
					 lower=ests[1],
					 upper=ests[3])
	all_coefs[[i]] = tdt
}
all_coefs_dt = do.call(rbind, all_coefs)
ggplot(all_coefs_dt) + geom_point(aes(y=model_nm, x=med)) +
					   geom_errorbarh(aes(y=model_nm, 
					   					  xmin=lower, 
					   					  xmax=upper),
					   					height=0.1) +
					   geom_vline(aes(xintercept=0), linetype="dashed") +
					   theme_classic() + ylab("Conditional independency") + xlab("Relationship value")
#ggsave("test_dag.pdf", width=4, height=3)

# Compare the b1 coefficients across models
# pdf("test_dag.pdf", width=4, height=3)
# plot(coeftab(test1, test2, test3), pars="b1", prob=0.95, main="", xlab="")
# dev.off()


```

Check some basic model assumptions

```{r}

# Median residuals
pred2 = link(tadmod2)
mean_pred = colMeans(pred2)
resid = svl_data$svl_z - mean_pred

# QQplot to check normality
df = data.frame(resid=resid, x=mean_pred)
p = ggplot(df, aes(sample = resid))
pq = p + stat_qq() + stat_qq_line() + theme_classic()

# Residual plot to check equal variance
pr = ggplot(df, aes(x=x, y=resid)) + geom_point() + geom_hline(aes(yintercept=0), linetype="dashed") + xlab("fitted value") + ylab("residual") + theme_classic()
pq + pr
ggsave("resid_plots_full_model.pdf", width=7, height=3)


```

Build and simulate DAGs

```{r}

taddag <- dagitty("dag{
Area -> Density [beta = 1]
Density -> SVL
Area -> Algae -> Density
Nutr -> Algae -> SVL
Nutr -> SVL
}")

coordinates(taddag) <- list(x=c(Area=1, Algae=5, Density=5,
                              Nutr = 1, SVL=10),
                         y=c(Area=5, Algae=5, Density=9,
                         	Nutr = 1, SVL=5))

rethinking::drawdag(taddag)

# Simulate the causal model specified above
n = 10000 # Number of samples
nutr = rnorm(n) # Independent 
area = rnorm(n) # Independent
algae = rnorm(n, nutr + area)
density = rnorm(n, -area + algae)
svl = rnorm(n, -density + 0.5*algae - nutr)

dat = data.frame(density=density,
				 nutr=nutr,
				 area=area,
				 algae=algae,
				 svl=svl)
head(dat, n=3)

# Simulation
n = 100
nutr = rnorm(n)
area = rnorm(n)
algae = seq(-2, 2, len=n) # Fix algae
density = rnorm(n, -area + algae)
svl = rnorm(n, -density + 0.5*algae - nutr)

dat = data.frame(density=density,
         nutr=nutr,
         area=area,
         algae=algae,
         svl=svl)

p1 = ggplot(dat) + geom_point(aes(x=algae, y=svl)) + theme_classic()
p1
#ggsave("counterfactual_algae.pdf", width=4, height=3)

# Simulation
n = 100
nutr = rnorm(n)
area = rnorm(n)
algae = seq(-2, 2, len=n) # Fix algae
density = rep(0, n)
svl = rnorm(n, -density + 0.5*algae - nutr)

dat = data.frame(density=density,
         nutr=nutr,
         area=area,
         algae=algae,
         svl=svl)

p2 = ggplot(dat) + geom_point(aes(x=algae, y=svl)) + theme_classic()
#ggsave("counterfactual_algae_density.pdf", width=4, height=3)
```

Estimate the full causal model

```{r}

# Fit the full causal model to estimate algae 
# manipulation
full_mod = quap(
	alist(
		# Algae -> SVL <- nutr, <- density
		svl ~ dnorm(mu1, sigma1),
		mu1 <- a0 + a_a*algae + a_d*density + a_n*nutr,
		a0 ~ dnorm(0, 5),
		a_a ~ dnorm(0, 2),
		a_d ~ dnorm(0, 2),
		a_n ~ dnorm(0, 2),
		sigma1 ~ dexp(1),
		# Algae -> density
		density ~ dnorm(mu2, sigma2),
		mu2 <- b0 + b_a*algae,
		b0 ~ dnorm(0, 5),
		b_a ~ dnorm(0, 2),
		sigma2 ~ dexp(1),
		# Nutrient -> SVL
		nutr ~ dnorm(mu3, sigma3),
		mu3 ~ dnorm(0, 5),
		sigma3 ~ dexp(1)
		), data=svl_data_z)

algae = seq(-2, 2, len=30)
sim_data = data.frame(algae=algae)
s = sim(full_mod, data=sim_data, vars=c("nutr", "density", "svl"), n=10000)



med = apply(s$svl, 2, quantile, 0.5)
lower = apply(s$svl, 2, quantile, 0.025)
upper = apply(s$svl, 2, quantile, 0.975)
s_dt = data.frame(algae=algae, med=med, 
				  lower=lower, upper=upper)

ggplot(s_dt) + geom_line(aes(x=algae, y=med)) +
			   geom_ribbon(aes(x=algae, ymin=lower, 
			   				   ymax=upper), alpha=0.2) + theme_classic() +
			   xlab("Algae") + ylab("SVL")
# ggsave("algae_counterfactual.pdf", width=4, height=3)



```

# Simulate tadpole data used in this example

```{r}


# Generate tadpole data
set.seed(25)
n = 25
nutr = rnorm(n)
area = rnorm(n)
algae = rnorm(n, nutr + area)
density = rnorm(n, -area + algae)
svl = rnorm(n, -density + 0.5*algae - nutr)

# Scale biologically
dat = data.frame(density=10 + 2*density,
				 nutr=0.002 + 0.0005*nutr,
				 area=10 + 2*area,
				 algae=3*algae,
				 svl=65 + 15*svl)

write.csv(dat, "tadpole_svl.csv", row.names=FALSE)

```





