---
title: 'Lecture code: Bayesian multivariate regression, Part II'
author: "Mark Wilber"
date: "`r Sys.Date()`"
output: html_document
---

Load in the dragonfly data

```{r}
library(ggplot2)

dfly_data = read.csv("dragonfly.csv")
dfly_data
ggplot(dfly_data, aes(x=bodylength, 
					  y=wingspan, 
					  color=population)) + 
					geom_point() + geom_line() +
					theme_classic() + xlab("Body length (mm)") +
					ylab("Wingspan (mm)")

#ggsave("dragonfly_plot.pdf", width=4, height=3)
```

Make a DAG

```{r}
library(dagitty)

dag <- dagitty("dag{
Population -> Wingspan
Bodysize -> Wingspan
}")

coordinates(dag) <- list(x=c(Population=0, 
							 Wingspan=2,
							 Bodysize=0),
                         y=c(Population=0,
                         	 Wingspan=-1,
                         	 Bodysize=-2))


pdf("dragon_dag.pdf", width=4, height=3)
rethinking::drawdag(dag)
dev.off()


```

```{r}
library(rethinking)
library(ggplot2)

# Simulate data to explore the effects of population on body size estimates

# set.seed(12)

# Strength of the effects
popeffects = seq(0, 1, len=10)
all_ests = list()

count = 1
for(peffect in popeffects){

	set.seed(10)
	n = 100
	population = sample(c(1, 2, 3), size=n, replace=TRUE)
	X = model.matrix(formula(~as.factor(population) - 1))
	pop_effects = c(0, 0, 0) # No effect of population on bodysize
	bodysize = rnorm(n, mean=X %*% pop_effects, sd=1)
	wingspan = rnorm(n, mean=peffect*2*X[, 1] - peffect*2*X[, 2] + peffect*0*X[, 3] + 
					         bodysize, sd=0.3)

	p1 = ggplot() + geom_point(aes(x=bodysize, y=wingspan, color=as.factor(population))) +
					theme_classic() + xlab("Body size, z") + ylab("Wingspan, z") +
	        guides(color=guide_legend(title="Pop"))
  
	# Just save no effect and strong effect
	if(peffect %in% c(0, 1)){
		ggsave(paste0("example_pop_effect", peffect, ".pdf"), width=4, height=3)
	}

	# Build dataframe to fit models
	tdat = data.frame(wingspan=wingspan, bodysize=bodysize)
	fit = quap(
			alist(
			   	wingspan ~ dnorm(mu, sigma),
			   	mu <- beta0 + beta1*(bodysize - mean(bodysize)),
			   	beta0 ~ dnorm(0, 1),
			   	beta1 ~ dnorm(0, 5),
			   	sigma ~ dexp(1)
			   	), data=tdat)

	ests = precis(fit, prob=0.95)
	td = data.frame(peffect=peffect, mean=ests$mean[2], 
					lower=ests$`2.5%`[2], upper=ests$`97.5%`[2])
	all_ests[[count]] = td
	count = count + 1
}

 # Plot the coefficient estimates
all_ests_df = do.call(rbind, all_ests)

ggplot(all_ests_df) + geom_point(aes(x=peffect, y=mean)) +
					  geom_errorbar(aes(x=peffect, ymin=lower, ymax=upper), width=0.05) +
					  geom_hline(aes(yintercept=1, color="True value"), linetype="dashed") +
					  scale_color_manual(values=c("black")) +
					  xlab("Effect of population") + ylab("Body size slope") +
					  theme_classic()
ggsave("slope_ests.pdf", width=6, height=3)


```

## Fit the model with discrete covariates

### Set-to-zero constraint

```{r}
head(dfly_data)

# Get design matrix
X = model.matrix(formula(~population), data=dfly_data)
X

df1 = data.frame(X)
colnames(df1) = c("int", "cat", "nav")
df1$wingspan = dfly_data$wingspan
mean(df1$wingspan)

# Fit the Bayesian model
fit1 = quap(
		alist(
			wingspan ~ dnorm(mu, sigma),
			mu <- beta0*int + beta1*cat + beta2*nav,
			beta0 ~ dnorm(11.85, 5),
			c(beta1, beta2) ~ dnorm(0, 5),
			sigma ~ dexp(1)), data=df1)
precis(fit1, prob=0.95)

# Extract the mean at Navarra
post = extract.samples(fit1, n=10000)
navarra = post[,1] + post[, 3]
quantile(navarra, c(0.025, 0.5, 0.975))

# Compare with a frequentist approach
summary(lm(wingspan ~ population, data=dfly_data))
```

### Means model

```{r}
X = model.matrix(formula(~population - 1), data=dfly_data)

df2 = data.frame(X)
colnames(df2) = c("arg", "cat", "nav")
df2$wingspan = dfly_data$wingspan
mean(df2$wingspan)

fit2 = quap(
		alist(
			wingspan ~ dnorm(mu, sigma),
			mu <- beta0*arg + beta1*cat + beta2*nav,
			c(beta0, beta1, beta2) ~ dnorm(11.85, 3),
			sigma ~ dexp(1)), data=df2)
precis(fit2, prob=0.95)

summary(lm(wingspan ~ population - 1, data=dfly_data))
```

### Fit the means model with population indices

```{r}

df3 = dfly_data[, c("wingspan", "population")]

# A helpful rethinking function to convert a factor into integer levels
df3$population2 = coerce_index(as.factor(df3$population))

fit3 = quap(
		alist(
			wingspan ~ dnorm(mu, sigma),
			mu <- beta[population2],
			beta[population2] ~ dnorm(11.85, 3),
			sigma ~ dexp(1)), data=df3)
precis(fit3, prob=0.95, depth=2)

```

Compare the posterior distributions to ensure we are getting the same answers
with the different ways of fitting the models

```{r}
# Compare the posterior distributions to ensure we are getting the same answers
post1 = extract.samples(fit1, n=10000)
post2 = extract.samples(fit2, n=10000)
post3 = extract.samples(fit3, n=10000, "beta")


cat1 = post1$beta0 + post1$beta1
cat2 = post2$beta1
cat3 = post3$beta[, 2]

# Plot Catalonia posteriors
ggplot() + geom_density(aes(x=cat1, color="Set-to-zero")) + 
		   geom_density(aes(x=cat2, color="Means model 1")) + 
		   geom_density(aes(x=cat3, color="Means model 2")) +
		   theme_classic() + xlab("Posterior wingspan, Catalonia")

ggsave("compare_posterior_cat.pdf", width=5, height=3)

# Plot aragon posteriors

arg1 = post1$beta0
arg2 = post2$beta0
arg3 = post3$beta[, 1]

ggplot() + geom_density(aes(x=arg1, color="Set-to-zero")) + 
		   geom_density(aes(x=arg2, color="Means model 1")) + 
		   geom_density(aes(x=arg3, color="Means model 2")) +
		   theme_classic() + xlab("Posterior wingspan, Aragon")

ggsave("compare_posterior_arg.pdf", width=5, height=3)

# Plot Navarro posteriors

nav1 = post1$beta0 + post1$beta2
nav2 = post2$beta2
nav3 = post3$beta[, 3]

ggplot() + geom_density(aes(x=nav1, color="Set-to-zero")) + 
		   geom_density(aes(x=nav2, color="Means model 1")) + 
		   geom_density(aes(x=nav3, color="Means model 2")) +
		   theme_classic() + xlab("Posterior wingspan, Catalonia")

ggsave("compare_posterior_nav.pdf", width=5, height=3)

```

Fit the model.  Explore prior predictive simulations first

```{r}
library(data.table)

dfly_data$population2 = coerce_index(as.factor(dfly_data$population))
dfly_data$bodylength_z = scale(dfly_data$bodylength)

# Prior simulations
n=100
bvals = seq(min(dfly_data$bodylength_z),
			      max(dfly_data$bodylength_z), len=n)

ints = rnorm(n, 11.85, 3)
slopes = rnorm(n, 0, 2)
preds = sapply(1:n, function(i) ints[i] + slopes[i]*bvals)

pred_dt = as.data.table(preds)
pred_dt$bvals = bvals
pred_dt_melt = melt(pred_dt, id.vars=c("bvals"))
ggplot(pred_dt_melt) + geom_line(aes(x=bvals, y=value, group=variable)) + theme_classic() + 
	xlab("Body length z") + ylab("Wingspan (mm)")
ggsave("prior_simulation.pdf", width=4, height=3)
```

```{r}
library(patchwork)

# Fit with means model
dfly_data$population2 = coerce_index(as.factor(dfly_data$population))
dfly_data$bodylength_z = scale(dfly_data$bodylength)

# Fit the model
mod1 = quap(
		alist(
			wingspan ~ dnorm(mu, sigma),
			mu <- alpha[population2] + beta1*bodylength_z,
			alpha[population2] ~ dnorm(11.85, 3),
			beta1 ~ dnorm(0, 2),
			sigma ~ dexp(1 / 2)),
		data=dfly_data)

# Residuals
# Median residuals
pred = link(mod1)
mean_pred = colMeans(pred)
resid = dfly_data$wingspan - mean_pred

# QQplot to check normality
df = data.frame(resid=resid, x=dfly_data$bodylength_z)
p = ggplot(df, aes(sample = resid))
pq = p + stat_qq() + stat_qq_line() + theme_classic()

# Residual plot to check equal variance
pr = ggplot(df, aes(x=x, y=resid)) + geom_point() + geom_hline(aes(yintercept=0), linetype="dashed") + xlab("bodylength_z") + ylab("residual") + theme_classic()
pq + pr
ggsave("residual_plot_dragon.pdf", width=6, height=3)


# Plot the mean predictions

# Make new data
bvals = seq(min(dfly_data$bodylength_z),
			max(dfly_data$bodylength_z), len=50)
newdata = expand.grid(bvals, c(1, 2, 3))
colnames(newdata) = c("bodylength_z", "population2")

# Simulate predictions from model
pred2 = data.frame(t(link(mod1, newdata, n=50)))
full_dat = data.table(cbind(newdata, pred2))
full_dat_melt = melt(full_dat, id.vars=c("bodylength_z", "population2"))
full_dat_melt$group = paste0(full_dat_melt$population2,
							 full_dat_melt$variable)

# Plot the observed vs predicted
ggplot(full_dat_melt) + 
	geom_line(aes(x=bodylength_z, y=value, group=group, color=as.factor(population2)), alpha=0.25) +
	geom_point(data=dfly_data, aes(x=bodylength_z, y=wingspan, color=as.factor(population2)), size=3) +
	geom_line(data=dfly_data, aes(x=bodylength_z, y=wingspan, color=as.factor(population2))) + 
	xlab("Body length z") + 
	ylab("Wingspan") + 
	labs(color=element_text("Population")) +
	theme_classic()
ggsave("obs_pred_dragonfly.pdf", width=5.5, height=3)


```

```{r}
# Plot coefficient estimates
post = extract.samples(mod1, n=10000)

# Is there a positive effect of body size?
quantile(post$beta1, c(0.025, 0.975))
mean(post$beta1 / sd(dfly_data$bodylength))

# Is Aragon different then Catalonia?
quantile(post$alpha[, 1] - post$alpha[, 2], c(0.025, 0.975))


```

## Multicollinearity

Understanding multicollinearity through simulation

```{r}
library(rethinking)

data(milk)
d = milk[, c("kcal.per.g", "perc.fat", "perc.lactose")]
head(d)
dim(d)

pdf("multicollinearity.pdf", width=5, height=4)
pairs(d)
dev.off()

cor(d)

```

Fit three different models to examine the effects of collinearity

```{r}

d_scale = data.frame(scale(d))
mod_lact = quap(
			alist(
				kcal.per.g ~ dnorm(mu, sigma),
				mu <- beta0 + beta1*perc.lactose,
				beta0 ~ dnorm(0, 3),
				beta1 ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=d_scale)
precis(mod_lact, prob=0.95)

mod_fat = quap(
			alist(
				kcal.per.g ~ dnorm(mu, sigma),
				mu <- beta0 + beta1*perc.fat,
				beta0 ~ dnorm(0, 3),
				beta1 ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=d_scale)
precis(mod_fat, prob=0.95)


mod_total = quap(
			alist(
				kcal.per.g ~ dnorm(mu, sigma),
				mu <- beta0 + beta1*perc.fat + beta2*perc.lactose,
				beta0 ~ dnorm(0, 3),
				c(beta1, beta2) ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=d_scale)
precis(mod_total, prob=0.95)

```


```{r}

# DAG 1
n = 100
tsd = 1 # Decrease to increase correlation
ff = rnorm(n)
fat = rnorm(n, -ff, sd=tsd)
lact = rnorm(n, ff, sd=tsd)
kilo = rnorm(n, fat - lact)

df = data.frame(k=kilo, f=fat, l=lact)

pairs(df)
cor(df)

sim_mod = quap(
			alist(
				k ~ dnorm(mu, sigma),
				mu <- beta0 + betaf*f + betal*l,
				beta0 ~ dnorm(0, 3),
				c(betaf, betal) ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=df)
precis(sim_mod, prob=0.95)
```

```{r}

# DAG 1
n = 100
tsd = 0.5 # Decrease to increase correlation
ff = rnorm(n)
fat = rnorm(n, -ff, sd=tsd)
lact = rnorm(n, ff, sd=tsd)
kilo = rnorm(n, fat - lact)

df = data.frame(k=kilo, f=fat, l=lact)

pairs(df)
cor(df)

sim_mod = quap(
			alist(
				k ~ dnorm(mu, sigma),
				mu <- beta0 + betaf*f + betal*l,
				beta0 ~ dnorm(0, 3),
				c(betaf, betal) ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=df)
precis(sim_mod, prob=0.95)
```

```{r}

# DAG 1
n = 100
tsd = 0.1 # Decrease to increase correlation
ff = rnorm(n)
fat = rnorm(n, -ff, sd=tsd)
lact = rnorm(n, ff, sd=tsd)
kilo = rnorm(n, fat - lact)

df = data.frame(k=kilo, f=fat, l=lact)

pairs(df)
cor(df)

sim_mod = quap(
			alist(
				k ~ dnorm(mu, sigma),
				mu <- beta0 + betaf*f + betal*l,
				beta0 ~ dnorm(0, 3),
				c(betaf, betal) ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=df)
precis(sim_mod, prob=0.95)
```

Explore the posteriors

```{r}
library(patchwork)
library(ggplot2)
library(rethinking)

sds = c(1, 0.5, 0.1, 0.01)
diffs = list()

count = 1
for(tsd in sds){

	# DAG 1
	set.seed(10)
	n = 100
	ff = rnorm(n)
	fat = rnorm(n, -ff, sd=tsd)
	lact = rnorm(n, ff, sd=tsd)
	kilo = rnorm(n, fat - lact)

	df = data.frame(k=kilo, f=fat, l=lact)

	sim_mod = quap(
				alist(
					k ~ dnorm(mu, sigma),
					mu <- beta0 + betaf*f + betal*l,
					beta0 ~ dnorm(0, 3),
					c(betaf, betal) ~ dnorm(0, 3),
					sigma ~ dexp(1)
					), data=df)

	post = extract.samples(sim_mod)
	prior = rnorm(10000, 0, 3)
	
	diffs[[count]] = post$betaf - post$betal
	count = count + 1

	p1 = ggplot() + geom_point(aes(x=post$betal, y=post$betaf)) + theme_classic() +
			   xlab("beta_l") + ylab("beta_f")
	p2 = ggplot() + geom_density(aes(x=post$betal, color="post.")) + 
					geom_density(aes(x=prior, color="prior")) + 
					theme_classic() + xlab("beta_l")
	p3 = ggplot() + geom_density(aes(x=post$betaf, color="post.")) + 
					geom_density(aes(x=prior, color="prior")) +  theme_classic() + xlab("beta_f")

	ptot = p1 + p2 + p3
	ggsave(paste0("compare_posterior_sd=", tsd, ".pdf"), width=8, height=3)
}

# Plot the histograms of the differences
ggplot() + geom_density(aes(x=diffs[[1]], color="sd = 1 (Low correlation)")) +
           geom_density(aes(x=diffs[[2]], color="sd = 0.5 (Medium correlation)")) +
           geom_density(aes(x=diffs[[2]], color="sd = 0.1 (High correlation)")) +
           geom_density(aes(x=diffs[[3]], color="sd = 0.01 (Very high correlation)")) + theme_bw() + xlab("beta_fat - beta_lactose")

ggsave("differences_in_beta.pdf", width=8, height=4)

```

Demonstrate that the difference $\beta_f - \beta_l$ is identifiable even when parameters are highly correlated.

```{r}
# DAG 1
set.seed(10)
n = 100
tsd= 0.01 # High correlation between lactose and fat
ff = rnorm(n)
fat = rnorm(n, -ff, sd=tsd)
lact = rnorm(n, ff, sd=tsd)
kilo = rnorm(n, fat - lact)

df = data.frame(k=kilo, f=fat, l=lact)

sim_mod = quap(
  alist(
    k ~ dnorm(mu, sigma),
    mu <- beta0 + betaf*f + betal*l,
    beta0 ~ dnorm(0, 3),
    c(betaf, betal) ~ dnorm(0, 3),
    sigma ~ dexp(1)
  ), data=df)

# Plot the posterior of the difference
post = extract.samples(sim_mod)
tdiff = post[, 2] - post[, 3]
ggplot() + geom_density(aes(x=tdiff)) + xlab("beta_f - beta_l") + ylab("density") + theme_bw()
precis(tdiff)
```


Same as above, but with a lot of samples

```{r}
library(patchwork)
library(ggplot2)
library(rethinking)

sds = c(1, 0.5, 0.1, 0.01)
diffs = list()

count = 1
for(tsd in sds){

	# DAG 1
	set.seed(10)
	n = 10000
	ff = rnorm(n)
	fat = rnorm(n, -ff, sd=tsd)
	lact = rnorm(n, ff, sd=tsd)
	kilo = rnorm(n, fat - lact)

	df = data.frame(k=kilo, f=fat, l=lact)

	sim_mod = quap(
				alist(
					k ~ dnorm(mu, sigma),
					mu <- beta0 + betaf*f + betal*l,
					beta0 ~ dnorm(0, 3),
					c(betaf, betal) ~ dnorm(0, 3),
					sigma ~ dexp(1)
					), data=df)

	post = extract.samples(sim_mod)
	prior = rnorm(10000, 0, 3)

	diffs[[count]] = post$betal + post$betaf
	count = count + 1

	p1 = ggplot() + geom_point(aes(x=post$betal, y=post$betaf)) + theme_classic() +
			   xlab("beta_l") + ylab("beta_f")
	p2 = ggplot() + geom_density(aes(x=post$betal, color="post.")) + 
					geom_density(aes(x=prior, color="prior")) + 
					theme_classic() + xlab("beta_l")
	p3 = ggplot() + geom_density(aes(x=post$betaf, color="post.")) + 
					geom_density(aes(x=prior, color="prior")) +  theme_classic() + xlab("beta_f")

	ptot = p1 + p2 + p3
	ggsave(paste0("compare_posterior_sd=", tsd, "big_sample.pdf"), width=8, height=3)
}

```






```{r}
df = read.csv("milk_challenge.csv")
df


mod_fat = quap(
			alist(
				kilo ~ dnorm(mu, sigma),
				mu <- beta0 + betaf*fat,
				beta0 ~ dnorm(0, 3),
				betaf ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=df)
precis(mod_fat, prob=0.95)



mod_fat_lac = quap(
			alist(
				kilo ~ dnorm(mu, sigma),
				mu <- beta0 + betaf*fat + betal*lactose,
				beta0 ~ dnorm(0, 3),
				betaf ~ dnorm(0, 3),
				betal ~ dnorm(0, 3),
				sigma ~ dexp(1)
				), data=df)
precis(mod_fat_lac, prob=0.95)

# You cannot approximate the total effect of lactose and fat on energy content because of co-linearity. 

```

## Milk challenge

```{r}
df = read.csv("milk_challenge.csv")
df

# Fit a model with only fat
mod_fat = quap(
 			alist(
 				kilo ~ dnorm(mu, sigma),
 				mu <- beta0 + betaf*fat,
 				beta0 ~ dnorm(0, 3),
 				betaf ~ dnorm(0, 3),
 				sigma ~ dexp(1)
 				), data=df)

precis(mod_fat, prob=0.95)

# Fit a model with both covariates
mod_fat_lact = quap(
 			alist(
 				kilo ~ dnorm(mu, sigma),
 				mu <- beta0 + betaf*fat + betal*lactose,
 				beta0 ~ dnorm(0, 3),
 				c(betaf, betal) ~ dnorm(0, 3),
 				sigma ~ dexp(1)
 				), data=df)

precis(mod_fat_lact, prob=0.95)
```