---
title: "final_project_code"
author: "Sarah Harkins"
date: "`r Sys.Date()`"
output: html_document
---


```{r}
library(dagitty)

# Build your DAG
dag <- dagitty("dag{
temperature -> flowers.visited 
insect.RTU -> flowers.visited 
temperature -> floral.density -> flowers.visited 
net.treatment -> flowers.visited 
}")

coordinates(dag) <- list(x=c(floral.density=2, temperature=1, insect.RTU=3, net.treatment= 2,
                             flowers.visited=2),
                         y=c(floral.density=2, temperature=1, insect.RTU=1, net.treatment= -2 ,
                             flowers.visited=0))
# Visualize DAG
rethinking::drawdag(dag)
```


```{r}
library(ggplot2)
data = read.csv("MNP_pollination_Larrea_Malacothrix.csv")
# data
head(data)

#ggplot(data) + geom_point(aes(x=floral.density, y=flowers.visited)) + theme_classic() + xlab("floral density") + ylab("flowers visited")


# ggplot(data) + geom_point(aes(x=net.treatment, y=flowers.visited)) + theme_classic() + xlab("net treatment") + ylab("flowers visited")


#ggplot(data) + geom_point(aes(x=insect.RTU, y=flowers.visited)) + theme_classic() + xlab("insect RTU") + ylab("flowers visited")
```

We are going to consider "flowers.visited" as our response variable to judge the influence of the facilitative relatioinship.
```{r}
library(ggplot2)
library(rethinking)
library(patchwork)
library(data.table)

# Normalizing the data 
data$floral_density_z = as.vector((scale(data$floral.density)))
data$temp_z = as.vector(scale(data$temperature))

# converting the treatment type and insect.RTU into factors 
data$treatment_id = coerce_index(as.factor(data$net.treatment)) 
data$insect_id = coerce_index(as.factor(data$insect.RTU)) 

head(data)

```


```{r}

data = data[!is.na(data$flowers.visited), ]
#data = data[!is.na(data$floral_density_z), ]

data
#data[complete.cases(data), ]

#data_clean <- na.omit(data)
#data_clean

data$flowersvisited = data$flowers.visited
data
```

```{r}
mean(data$flowersvisited)
#mean(data$flowers.visited)
#colMeans(data, dims=1)
```

# Model 1
This model is an extremely basic means model wth weak regularization in it's priors. 


```{r}
mod_1 = alist(
		flowersvisited ~ dpois(mu), # response is flowers visited (flowers visited is un-normalized in data frame)
		log(mu) <- alpha[treatment_id],
		alpha[treatment_id] ~ dnorm(0, 3)
		)

# Fit with ulam
stan_data_1 = data[, c("treatment_id", "flowersvisited")]

mod_1_fit = ulam(mod_1, data=stan_data_1, iter=2000,
				warmup=500, chains=4, cores=4, log_lik=TRUE)

mod_1_fit = ulam(mod_1, data=stan_data_1, iter=2000,
				warmup=500, chains=4, cores=4)

#precis(mod_1_fit)
precis(mod_1_fit, depth = 2)
```



# Checking Model 1's Fit

```{r}
traceplot(mod_1_fit)


# Check model fit
postcheck(mod_1_fit)
```

## Explore posterior predictions

Look at model predictions (alternative approach to postcheck)

```{r}
mod1_pred = link(mod_1_fit) # Get predictions (outputs the systematic component )
mpv =  as.vector(mod1_pred) 
mod1_pred_counts = array(rpois(length(mpv), mpv), dim=dim(mod1_pred)) # Draw poisson samples

p1 = bayesplot::ppc_dens_overlay(y=log(data$flowersvisited + 1), 
								 yrep=log(mod1_pred_counts[1:500, ] + 1))

p1 + theme_classic() + xlab("log(flowersvisied + 1)") + ylab("Density")
```


# Model 2
```{r}
mod_2 = alist(
		flowersvisited ~ dpois(mu), # response is flowers visited (flowers visited is un-normalized in data frame)
		log(mu) <- beta1 + alpha[treatment_id],
		beta1 ~ dnorm(2, 1),
		alpha[treatment_id] ~ dnorm(0, 3)
		)

# Fit with ulam
stan_data_2 = data[, c("treatment_id", "flowersvisited")]

mod_2_fit = ulam(mod_2, data=stan_data_2, iter=2000,
				warmup=500, chains=4, cores=4, log_lik=TRUE)
```


# Comparing Model 1 and Model 2
```{r}
compare(mod_1_fit, mod_2_fit, func=PSIS)
```

Considering the interval generated by $\text{dPSIS} \pm 2 \text{dSE}$, we get the following interval [0.32, 1.08]. Since $0 \notin [0.32, 1.08]$, we cna conclude that Model 2 is predictively superior to Model 1.  


## Explore posterior predictions

Look at model predictions (alternative approach to postcheck)

```{r}
mod2_pred = link(mod_2_fit) # Get predictions (outputs the systematic component )
mpv =  as.vector(mod2_pred) 
mod2_pred_counts = array(rpois(length(mpv), mpv), dim=dim(mod2_pred)) # Draw poisson samples

p1 = bayesplot::ppc_dens_overlay(y=log(data$flowersvisited + 1), 
								 yrep=log(mod2_pred_counts[1:500, ] + 1))

p1 + theme_classic() + xlab("log(flowersvisied + 1)") + ylab("Density")
```



# Model 3
```{r}
data_v2 = data[!is.na(data$floral_density_z), ]

stan_data_3 = data_v2[, c("treatment_id", "flowersvisited", "floral_density_z")]

mod_3 = alist(
		flowersvisited ~ dpois(mu), # response is flowers visited (flowers.visited is unnormalized in dataframe)
		log(mu) <-  beta1 + alpha[treatment_id] + beta_density*floral_density_z,
		beta1 ~ dnorm(2, 1), 
		alpha[treatment_id] ~ dnorm(0, 1), # variation in # of flowers visited for each treatment 
		c(beta_density) ~ dnorm(0, 1) # effect of floral density on flower visitation 
		)

mod_3_fit = ulam(mod_3, data=stan_data_3, iter=2000,
				warmup=500, chains=4, cores=4, log_lik=TRUE)

precis(mod_3_fit, depth = 2)


```
# Comparing Model 1 and Model 3
```{r}
compare(mod_1_fit, mod_3_fit, func=PSIS)
```



## Explore posterior predictions

Look at model predictions (alternative approach to postcheck)

```{r}
mod3_pred = link(mod_3_fit) # Get predictions (outputs the systematic component )
mpv =  as.vector(mod3_pred) 
mod3_pred_counts = array(rpois(length(mpv), mpv), dim=dim(mod3_pred)) # Draw poisson samples
dim(mod3_pred_counts)
dim(log(mod3_pred_counts[1:500, ] + 1))


p1 = bayesplot::ppc_dens_overlay(y=log(data$flowersvisited + 1), 
								 yrep=log(mod3_pred_counts[1:500, ] + 1))

p1 + theme_classic() + xlab("log(flowersvisied + 1)") + ylab("Density")
```



# Model 4 - Model with insect as a fixed effect 
```{r}
#mod_4 = alist(
#		flowersvisited ~ dpois(mu), # response is flowers visited (flowers.visited is unnormalized in dataframe)
#		log(mu) <- beta1 + alpha[treatment_id] + beta_density*floral_density_z + beta_insect*insect_id,
#		beta1 ~ dnorm(2, 1), # 2 is the "grand" mean of flowers.visited 
#		alpha[treatment_id] ~ dnorm(0, sigma_treatment), # variation in # of flowers visited for each treatment 
#		sigma_treatment ~ dexp(1),
#		c(beta_density, beta_insect) ~ dnorm(0, 2) 
#		)

mod_4 = alist(
		flowersvisited ~ dpois(mu), # response is flowers visited (flowers.visited is unnormalized in dataframe)
		log(mu) <- beta1 + alpha[treatment_id] + beta_density*floral_density_z + beta_insect*insect_id,
		beta1 ~ dnorm(2, 1), # 2 is the "grand" mean of flowers.visited 
		alpha[treatment_id] ~ dnorm(0, 1), # variation in # of flowers visited for each treatment 
		c(beta_density, beta_insect) ~ dnorm(0, 2) 
		)

stan_data_4 = data_v2[, c("treatment_id", "flowersvisited", "floral_density_z", "insect_id")]

mod_4_fit = ulam(mod_4, data=stan_data_4, iter=2000,
				warmup=500, chains=4, cores=4, log_lik=TRUE)

precis(mod_4_fit, depth = 2)
```


# Comparing Models
```{r}
compare(mod_1_fit, mod_2_fit, mod_3_fit, mod_4_fit, func=PSIS)
```





# Model 5
```{r}
# Partial Pooling Model 

mod1 = alist(
		flowersvisited ~ dpois(mu),
		log(mu) <- beta1 + alpha[treatment_id] + beta_temp*temp_z + beta_insect*insect_id + beta_floraldensity*floral_density_z,
		beta1 ~ dnorm(0.5, 1),
		alpha[treatment_id] ~ dnorm(0, sigma_treat),
		c(beta_temp, beta_insect, beta_floraldensity) ~ dnorm(0, 1),
		sigma_treat ~ dhalfnorm(0, 1)
		)


# Fit with ulam
stan_data = data_clean[, c("treatment_id", "flowersvisited", "insect_id", "floral_density_z")]
mod1_fit = ulam(mod1, data=stan_data, iter=2000,
				warmup=500, chains=4, cores=4, log_lik=TRUE)

precis(mod1_fit)
```


```{r}
traceplot(fox_mod1_fit)
trankplot(fox_mod1_fit)
precis(fox_mod1_fit)
```

Here is Bayesian model with some regularization which helps the fitting

```{r}
gt_dat$location_id = coerce_index(as.factor(gt_dat$location))

mod2 = alist(
		count ~ dpois(mu),
		log(mu) <- alpha[location_id] + beta_forest*forest_z + beta_elevation*elevation_z + beta_fe*forest_z*elevation_z + beta_wind*wind_z,
		alpha[location_id] ~ dnorm(0, 3), # Weak regularization
		c(beta_forest, beta_elevation, beta_fe, beta_wind) ~ dnorm(0, 1)
		)

# Takes a little bit of time to fit
stan_data = gt_dat[, c("location_id", "count", 
						"forest_z", "elevation_z", 
						"wind_z")]
mod2_fit = ulam(mod2, data=stan_data, iter=5000,
				warmup=1000, chains=4, cores=4, log_lik=TRUE)

#precis(mod2_fit, depth=2)

#traceplot(mod2_fit, pars=c("beta_elevation", "beta_forest"))

# Check model fit
#postcheck(mod2_fit)
```


Fit no pooling model...use a frequentist approach for comparison with no regularization

```{r}
fit_nopool = glm(count ~ as.factor(location) + forest_z + elevation_z +
						  forest_z*elevation_z + wind_z - 1, 
						  family="poisson", data=gt_dat)
```




## Model comparison (also using loo for comparison)

```{r}
compare(mod1_fit, mod2_fit, func=PSIS)
```

```{r}
# Generate predictions for each site for hierarchical model
post1 = extract.samples(mod1_fit)
beta1 = post1$beta1
alpha = post1$alpha

# Get location posteriors for hierarchical model
location_mod1 = alpha + as.vector(beta1)
location_mod1_mean = colMeans(exp(location_mod1))

# Generate predictions for no pooling model
post2 = extract.samples(mod2_fit)
location_mod2 = exp(post2$alpha)
location_mod2_mean = colMeans(location_mod2)

# Frequentist predictions
# location_mod2_mean = exp(coef(fit_nopool)[1:267])

# Choose some points to examine
ind = 75:100
ggplot() + geom_point(aes(x=ind, 
						  y=location_mod1_mean[ind], color="Partial pooling")) +
		   geom_point(aes(x=ind, 
						  y=location_mod2_mean[ind], color="No pooling")) +
		   geom_hline(aes(yintercept=mean(exp(beta1))), linetype="dashed") + theme_classic() + xlab("Location number") +
		   ylab("Expected bird abundance")

ggsave("compare_partial_and_no_pooling.pdf", width=5, height=3)

```


