---
title: "Harkins_Final_Project"
author: "Sarah Harkins"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library(deSolve)
library(ggplot2)
library(bbmle)
library(MASS)
library(bbmle)
```

```{r}
# Load in the pathogen data
black_data = read.csv("/Users/harkins/Desktop/560/black_0.05_mg_per_liter.csv")
green_data = read.csv("/Users/harkins/Desktop/560/green_control.csv")
red_data = read.csv("/Users/harkins/Desktop/560/red_0.01_mg_per_liter.csv")

head(black_data)
head(red_data)
head(green_data)
```
```{r}
#combining data frames to one mega-data frame
observed_df = data.frame(black_data)

observed_df$red_data_day <- red_data$day
observed_df$red_data_parasite <- red_data$parasite

observed_df$green_data_day <- green_data$day
observed_df$green_data_parasite <- green_data$parasite

observed_df
```

```{r}
#install.packages("ggplot2") # If you don't already have ggplot
library(ggplot2)

ggplot(data=observed_df, aes(x=day, y=parasite, color="0.05 mg l^{-1}")) + geom_line() + 
            geom_point() + 
  geom_point(aes(x=red_data_day, y=red_data_parasite, color="0.01 mg l^{-1}")) +
            geom_line(aes(x=red_data_day, y=red_data_parasite, color="0.01 mg l^{-1}")) + 
  geom_point(aes(x=green_data_day, y=green_data_parasite, color="Control")) +
            geom_line(aes(x=green_data_day, y=green_data_parasite, color="Control"))

```



```{r}
pi_model_HT1 = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  beta = params["beta"]
  d = params["d"] #death term of immune particles
  p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  f = params["f"] #rate at which the immune particles are produce due to detected parasite 
  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: the parasite measure in mass 
  
  #considering a Holling Type I functional response
  dP = beta*P - p*I*P
  dI = B - d*I +f*P*I   

  #try to show delayed pred/prey wave between parasite and immune system 
  # capture one "wave" over the 30 day experiment 
  
  return(list(c(dP, dI)))
}
```

```{r}
pi_model_HT2 = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 
  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: the parasite measure in mass 
  
    #considering a Holling Type II function response 
  dP = beta*P - I*P*c/(a+P)
  dI = B - d*I + P*I*b/(a+P) 
 
  return(list(c(dP, dI)))
}
```

```{r}
pi_model_HT3 = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 

  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 
  
  #considering a Holling Type III function response 
  dP = beta*P - I*P^2*c/(a^2+P^2)
  dI = B - d*I + I*P^2*b/(a^2+P^2)

  return(list(c(dP, dI)))
}
```


```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT1 = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT1, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT2 = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT2, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT3 = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT3, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```


Finding optimal parameters for each system of ODEs then plotting the simulation with those parameters against the observed data from the paper. 

Fitting to the control data: 
```{r}
#using optim to find appropriate parameters
params_HT1 = c(beta=0.11, d = 0.2, p = 0.3 , B = 0.005, f=0.1)

#FOR HOLLINGS TYPE I
fit0_HT1 = optim(params_HT1, rss_pi_HT1, data=observed_df$green_data_parasite)
#fit0_HT1$par
#fit0$convergence
fit1_HT1 = optim(fit0_HT1$par, rss_pi_HT1, data=observed_df$green_data_parasite, control=list(maxit=10000))
#fit1_HT1$par
fit3_HT1 = optim(fit0_HT1$par, rss_pi_HT1, data=observed_df$green_data_parasite, control=list(maxit=10000))
fit3_HT1$par
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)

pi_results_min_HT1 = ode(init_vals, times, pi_model_HT1, fit3_HT1$par) # fit0_HT1$par

pi_results_min_df_HT1 = data.frame(pi_results_min_HT1)
#pi_results_min_df_HT1

ggplot() + geom_line(aes(x=pi_results_min_df_HT1$time, y=pi_results_min_df_HT1$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```


```{r}
params_HT2 = c(beta=0.1, d = 0.2, B = 0.005, a = 2.6, c = 0.7, b=0.1)


#FOR HOLLINGS TYPE II
fit0_HT2 = optim(params_HT2, rss_pi_HT2, data=observed_df$green_data_parasite, control=list(maxit=10000))
#fit0_HT2$par
#fit0$convergence
fit1_HT2 = optim(fit0_HT2$par, rss_pi_HT2, data=observed_df$green_data_parasite)
#fit1_HT2$par
fit3_HT2 = optim(fit1_HT2$par, rss_pi_HT2, data=observed_df$green_data_parasite)
fit3_HT2$par
#fit0_HT2$par
```

```{r}
#RSS of the Holling type II model 
rss_pi_HT2(fit0_HT2$par,observed_df$green_data_parasite )
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)

pi_results_min_HT2 = ode(init_vals, times, pi_model_HT2, fit3_HT2$par)

pi_results_min_df_HT2 = data.frame(pi_results_min_HT2)
#pi_results_min_df_HT2

ggplot() + geom_line(aes(x=pi_results_min_df_HT2$time, y=pi_results_min_df_HT2$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```

```{r}
params_HT3 = c(beta=0.1, d = 0.2 , B = 0.005, a = 2.6, c = 0.7, b=0.1)

#FOR HOLLINGS TYPE III
fit0_HT3  = optim(params_HT3, rss_pi_HT3, data=observed_df$green_data_parasite, method="L-BFGS-B", lower=c(d=0), upper=c(d=3))
#fit0_HT3$par
#fit0$convergence
fit1_HT3 = optim(fit0_HT3$par, rss_pi_HT3, data=observed_df$green_data_parasite)
#fit1_HT3$par
fit3_HT3 = optim(fit1_HT3$par, rss_pi_HT3, data=observed_df$green_data_parasite)
fit3_HT3$par
```

```{r}
#RSS of the Holling type III model 
rss_pi_HT3(fit3_HT3$par, observed_df$green_data_parasite)
```


```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)

pi_results_min_HT3 = ode(init_vals, times, pi_model_HT3, fit0_HT3$par)

pi_results_min_df_HT3 = data.frame(pi_results_min_HT3)
#pi_results_min_df_HT3

ggplot() + geom_line(aes(x=pi_results_min_df_HT3$time, y=pi_results_min_df_HT3$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```

Fitting to the data for the 0.01 mg l^-1 microplastics ("red" data): 

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT1_0.01 = function(params, data_frame){
  times = observed_df$red_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$red_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT1, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$red_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT2_0.01 = function(params, data_frame){
  times = observed_df$red_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$red_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT2, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$red_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT3_0.01 = function(params, data_frame){
  times = observed_df$red_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$red_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT3, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$red_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#using optim to find appropriate parameters
params_HT1_red = c(beta=0.11, d = 0.2, p = 0.3 , B = 0.005, f=0.1)

#FOR HOLLINGS TYPE I
fit0_HT1_red = optim(params_HT1_red, rss_pi_HT1_0.01, data=observed_df$red_data_parasite)
#fit0_HT1$par
#fit0$convergence
fit1_HT1_red = optim(fit0_HT1_red$par, rss_pi_HT1_0.01, data=observed_df$red_data_parasite, control=list(maxit=10000))
#fit1_HT1$par
fit3_HT1_red = optim(fit0_HT1_red$par, rss_pi_HT1_0.01, data=observed_df$red_data_parasite, control=list(maxit=10000))
fit3_HT1_red$par
```

```{r}
times_red = observed_df$red_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals_red = c(P= observed_df$red_data_parasite[1], I = 0.1)
#init_vals_red

pi_results_min_HT1_red = ode(init_vals_red, times_red, pi_model_HT1, fit3_HT1_red$par) 
#pi_results_min_HT1_red

pi_results_min_df_HT1_red = data.frame(pi_results_min_HT1_red)
#pi_results_min_df_HT1_red

ggplot() + geom_line(aes(x=pi_results_min_df_HT1_red$time, y=pi_results_min_df_HT1_red$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```


```{r}
params_HT2 = c(beta=0.1, d = 0.2, B = 0.005, a = 2.6, c = 0.7, b=0.1)


#FOR HOLLINGS TYPE II
fit0_HT2_red = optim(params_HT2, rss_pi_HT2_0.01, data=observed_df$red_data_parasite, control=list(maxit=10000))
#fit0_HT2$par
#fit0$convergence
fit1_HT2_red = optim(fit0_HT2_red$par, rss_pi_HT2_0.01, data=observed_df$red_data_parasite)
#fit1_HT2$par
fit3_HT2_red = optim(fit1_HT2_red$par, rss_pi_HT2_0.01, data=observed_df$red_data_parasite)
fit3_HT2_red$par
#fit0_HT2_red$par
```

```{r}
#RSS of the Holling type II model 
rss_pi_HT2_0.01(fit0_HT2_red$par,observed_df$red_data_parasite )
```

```{r}
times_red = observed_df$red_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals_red = c(P= observed_df$red_data_parasite[1], I = 0.1)

pi_results_min_HT2_red  = ode(init_vals_red, times_red, pi_model_HT2, fit3_HT2_red$par)

pi_results_min_df_HT2_red  = data.frame(pi_results_min_HT2_red )
#pi_results_min_df_HT2

ggplot() + geom_line(aes(x=pi_results_min_df_HT2_red$time, y=pi_results_min_df_HT2_red$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```



```{r}
params_HT3_red = c(beta=0.1, d = 0.2 , B = 0.005, a = 2.6, c = 0.7, b=0.1)

#FOR HOLLINGS TYPE III
fit0_HT3_red  = optim(params_HT3_red, rss_pi_HT3_0.01, data=observed_df$red_data_parasite, method="L-BFGS-B", lower=c(d=0), upper=c(d=3))
#fit0_HT3$par
#fit0$convergence
fit1_HT3_red = optim(fit0_HT3_red$par, rss_pi_HT3_0.01, data=observed_df$red_data_parasite)
#fit1_HT3$par
fit3_HT3_red = optim(fit1_HT3_red$par, rss_pi_HT3_0.01, data=observed_df$red_data_parasite)
fit3_HT3_red$par
```

```{r}
#RSS of the Holling type III model 
rss_pi_HT3_0.01(fit3_HT3_red$par, observed_df$red_data_parasite)
```


```{r}
times_red = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals_red = c(P= observed_df$red_data_parasite[1], I = 0.1)

pi_results_min_HT3_red = ode(init_vals_red, times_red, pi_model_HT3, fit3_HT3_red$par)

pi_results_min_df_HT3_red = data.frame(pi_results_min_HT3_red)
#pi_results_min_df_HT3

ggplot() + geom_line(aes(x=pi_results_min_df_HT3_red$time, y=pi_results_min_df_HT3_red$P, color='Optimized Parameters'))+
           geom_point(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```
Plot of the simulated and observed parasite counts for the control experiment and 0.01 mg l^-1 microplastics:

```{r}
ggplot() + geom_line(aes(x=pi_results_min_df_HT1_red$time, y=pi_results_min_df_HT1_red$P, color='0.01 - Optimized Parameters'))+
           geom_point(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data"))+
       geom_line(aes(x=pi_results_min_df_HT1$time, y=pi_results_min_df_HT1$P, color='Control - Optimized Parameters'))+
       geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
       geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Log of Mean Parasite Intensity") + xlab("Time (Days)")+ggtitle("Holling Type I Functional Response Model")
```


```{r}
ggplot() + geom_line(aes(x=pi_results_min_df_HT2_red$time, y=pi_results_min_df_HT2_red$P, color='0.01 - Optimized Parameters'))+
           geom_point(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data"))+
       geom_line(aes(x=pi_results_min_df_HT2$time, y=pi_results_min_df_HT2$P, color='Control - Optimized Parameters'))+
       geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
       geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Log of Mean Parasite Intensity") + xlab("Time (Days)")+ggtitle("Holling Type II Functional Response Model")
```


```{r}
ggplot() + geom_line(aes(x=pi_results_min_df_HT3_red$time, y=pi_results_min_df_HT3_red$P, color='0.01 - Optimized Parameters'))+
           geom_point(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$red_data_day, y=observed_df$red_data_parasite, color="0.01 - Experimental Obsv. Parasite Data"))+
       geom_line(aes(x=pi_results_min_df_HT3$time, y=pi_results_min_df_HT3$P, color='Control - Optimized Parameters'))+
       geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
       geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Log of Mean Parasite Intensity") + xlab("Time (Days)")+ggtitle("Holling Type III Functional Response Model")
```


Now, we will calculate AIC to determine which of these three models is best suited. We will consider normal measurement error. 

Calculating AIC for the control data models: 
```{r}
HT1_nll = function(beta, d, p, B, f, sigma){
  
  # Setting initial conditions and parameters just like before
  times = observed_df$green_data_day
  params = c(beta=beta, d=d, p=p, B=B, f=f)
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT1, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  nll = -1*sum(dnorm(observed_df$green_data_parasite, mean=ode_res$P[1:length(times)], sd=sigma, log=TRUE)) #observed_df$green_data_parasite

  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit3_HT1$par)
#start_params


# Get the mle estimates for my parameters
fit_all_HT1 = mle2(HT1_nll, 
               start=list(beta = start_params[1], 
                          d = start_params[2], 
                          p = start_params[3],
                          B = start_params[4],
                          f = start_params[5],
                          sigma = 0.01),
               method="L-BFGS-B",
               lower=c(0, 0, 0, 0, 0, 0),
               upper=c(3, 3, 3, 3, 3, 3))


AIC(fit_all_HT1)
#AIC: -58.90567
```

```{r}

HT2_nll = function(beta, d, B, a, c, b, sigma){
  # Setting initial conditions and parameters just like before
  times = observed_df$green_data_day
  params = c(beta=beta, d=d, B=B, a=a, c=c, b=b)
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT2, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  nll = -1*sum(dnorm(observed_df$green_data_parasite, mean=ode_res$P[1:length(times)], sd=sigma, log=TRUE)) #observed_df$green_data_parasite

  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit3_HT2$par)
#start_params


# Get the mle estimates for my parameters
fit_all_HT2 = mle2(HT2_nll, 
               start=list(beta = start_params[1], 
                          d = start_params[2], 
                          B = start_params[3],
                          a = start_params[4],
                          c = start_params[5],
                          b = start_params[6],
                          sigma = 0.01),
               method="L-BFGS-B",
               lower=c(beta=0, d=0, B=0, a=0, c=0, b=0, sigma=0),
               upper=c(beta=3, d=2, B=3, a=3, c=3, b=3, sigma=3))
AIC(fit_all_HT2)
#AIC: -54.45174
```


```{r}

HT3_nll = function(beta, d, B, a, c, b, sigma){
  
  # Setting initial conditions and parameters just like before
  times = observed_df$green_data_day
  params = c(beta=beta, d=d, B=B, a=a, c=c, b=b)
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT3, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  nll = -1*sum(dnorm(observed_df$green_data_parasite, mean=ode_res$P[1:length(times)], sd=sigma, log=TRUE)) #observed_df$green_data_parasite

  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit3_HT3$par)
#start_params

HT3_nll(beta = start_params[1], 
                          d = start_params[2], 
                          B = start_params[3],
                          a = start_params[4],
                          c = start_params[5],
                          b = start_params[6],
                          sigma = 0.1)
# Get the mle estimates for my parameters
fit_all_HT3 = mle2(HT3_nll, 
               start=list(beta = start_params[1], 
                          d = 0.099, 
                          B = start_params[3],
                          a = start_params[4],
                          c = start_params[5],
                          b = start_params[6],
                          sigma = 0.1),
               method="L-BFGS-B",
               lower=c(beta=0, d=0, B=0, a=0, c=0, b=0, sigma=0),
               upper=c(beta=3, d=3, B=3, a=3, c=3, b=3, sigma=1))
#fit_all_HT3
AIC(fit_all_HT3)
#AIC: -56.41178
```




Calculating AIC for the 0.01 mg l^-1 microplastics data models: 

```{r}
HT1_nll_0.01 = function(beta, d, p, B, f, sigma){
  
  # Setting initial conditions and parameters just like before
  times = observed_df$red_data_day
  params = c(beta=beta, d=d, p=p, B=B, f=f)
  init_vals = c(P= observed_df$red_data_parasite[1], I = 0.1)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT1, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  nll = -1*sum(dnorm(observed_df$red_data_parasite, mean=ode_res$P[1:length(times)], sd=sigma, log=TRUE)) 

  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit0_HT1_red$par) #fit3_HT1_red$par
#start_params


# Get the mle estimates for my parameters
fit_all_HT1_0.01 = mle2(HT1_nll_0.01, 
               start=list(beta = start_params[1], 
                          d = start_params[2], 
                          p = start_params[3],
                          B = start_params[4],
                          f = start_params[5],
                          sigma = 0.1),
               method="L-BFGS-B",
               lower=c(beta=0.01, d=0.01, p=0.01, B=0.01, f=0.01, sigma=0.01),
               upper=c(beta=3, d=3, p=3, B=3, f=3, sigma=1))


AIC(fit_all_HT1_0.01)
#AIC: -31.29976 
```

```{r}

HT2_nll_0.01 = function(beta, d, B, a, c, b, sigma){
  # Setting initial conditions and parameters just like before
  times = observed_df$red_data_day
  params = c(beta=beta, d=d, B=B, a=a, c=c, b=b)
  init_vals = c(P= observed_df$red_data_parasite[1], I = 0.1)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT2, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  nll = -1*sum(dnorm(observed_df$red_data_parasite, mean=ode_res$P[1:length(times)], sd=sigma, log=TRUE)) #observed_df$green_data_parasite

  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit3_HT2_red$par) 
start_params


# Get the mle estimates for my parameters
fit_all_HT2_0.01 = mle2(HT2_nll_0.01, 
               start=list(beta = start_params[1], 
                          d = start_params[2], 
                          B = start_params[3],
                          a = start_params[4],
                          c = start_params[5],
                          b = start_params[6],
                          sigma = 0.01),
               method="L-BFGS-B",
               lower=c(beta=0, d=0, B=0, a=0, c=0, b=0, sigma=0),
               upper=c(beta=3, d=2, B=10, a=Inf, c=3, b=Inf, sigma=1))
AIC(fit_all_HT2_0.01)
#AIC: -29.03474
```


```{r}

HT3_nll_0.01 = function(beta, d, B, a, c, b, sigma){
  
  # Setting initial conditions and parameters just like before
  times = observed_df$red_data_day
  params = c(beta=beta, d=d, B=B, a=a, c=c, b=b)
  init_vals = c(P= observed_df$red_data_parasite[1], I = 0.1)
  
  # Simulate the model
  ode_res = as.data.frame(ode(func=pi_model_HT3, 
                                y=init_vals, 
                                times=times,
                                parms=params))
  nll = -1*sum(dnorm(observed_df$red_data_parasite, mean=ode_res$P[1:length(times)], sd=sigma, log=TRUE)) #observed_df$green_data_parasite

  return(nll)
  
}

# Test that the function is working
start_params = as.vector(fit0_HT3_red$par)
start_params

#HT3_nll(beta = start_params[1], 
#                          d = start_params[2], 
#                          B = start_params[3],
#                          a = start_params[4],
#                          c = start_params[5],
 #                         b = start_params[6],
 #                         sigma = 0.1)

# Get the mle estimates for my parameters
fit_all_HT3_0.01 = mle2(HT3_nll_0.01, 
               start=list(beta = start_params[1], 
                          d = 0.099, 
                          B = start_params[3],
                          a = start_params[4],
                          c = start_params[5],
                          b = start_params[6],
                          sigma = 0.1),
               method="L-BFGS-B",
               lower=c(beta=0.01, d=0.01, B=0.01, a=0.01, c=0.01, b=0.01, sigma=0),
               upper=c(beta=3, d=3, B=5, a=Inf, c=3, b=Inf, sigma=1))
#fit_all_HT3
AIC(fit_all_HT3_0.01)
#AIC: -27.67159
```





To prevent unbounded growth of the parasite, seen in the simulations of the Hollings Type I and II models, we will impose a carrying capacity via logistic growth in the growth term of the parasites equation. 

```{r}
pi_model_HT1_car_cap = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  #beta = params["beta"]
  d = params["d"] #death term of immune particles
  p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  K = params["K"] #carrying capacity of parasites in the tank 
  r = params["r"]
  f = params["f"]
  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 
  
  #considering a Holling Type I functional response
  dP = r*P*(1-P/K) - p*I*P
  dI = B + f*P*I - d*I   

  return(list(c(dP, dI)))
}
```


```{r}
pi_model_HT2_car_cap = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  #beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  K = params["K"] #carrying capacity of parasites in the tank 
  r = params["r"] #intrinsic growth rate of parasite 
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 

  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 
  
  #considering a Holling Type II function response 
  dP = r*P*(1-P/K) - I*P*c/(a+P)
  dI = B - d*I + P*I*b/(a+P)

  return(list(c(dP, dI)))
}
```

```{r}
pi_model_HT3_car_cap = function(time, state_vars, params){
  
  # Extract state variables
  P = state_vars['P'] #mass of parasites
  I = state_vars['I'] #immune particles 
  
  # Extract the parameters
  #beta = params["beta"]
  d = params["d"] #death term of immune particles
  #p = params["p"] #rate at which the immune particles kill parasite (predation of parasite by immune particles)
  B = params["B"] #constant background level of immune particles that are always present with or without disease/parasites present
  c = params["c"] #max per capita consumption rate 
  a = params["a"] #1/2 saturation rate
  K = params["K"] #carrying capacity of parasites in the tank 
  r = params["r"] #intrinsic growth rate of parasite 
  b = params["b"] #proportionality constant that "converts" detected parasites into more immune particles 

  #ODEs 
  #I: for immune system acting as the "predator" 
      #B is baseline level of immune particles
  #P: he parasite measure in mass 

  #considering a Holling Type III function response 
  dP = r*P*(1-P/K) - I*P^2*c/(a^2+P^2)
  dI = B - d*I + I*P^2*b/(a^2+P^2)
  


  return(list(c(dP, dI)))
}
```


```{r}
rss_pi_HT1_car_cap = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT1_car_cap, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT2_car_cap = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT2_car_cap, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```

```{r}
#function to compute the RSS for a given set of parameters 

rss_pi_HT3_car_cap = function(params, data_frame){
  times = observed_df$green_data_day #times of measurements from observed digitized data
  
  # Initial conditions
  init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)
  
  # Simulate the model.
  pred = ode(init_vals, times, pi_model_HT3_car_cap, params)
  pred_df = data.frame(pred)
  
  # Extract B from the model. Don't include the initial value because that is not in the data.
  predP = pred_df$P
  
  # Extract observed data
  obsP = observed_df$green_data_parasite
  
  # Compute residuals
  errors = obsP - predP
  
  rss = sum(errors^2)
  
  return(rss)
}
```


```{r}
#using optim to find appropriate parameters
params_HT1_car_cap = c(d = 0.2, p = 0.3 , B = 0.005, K=10, r=0.1, f=0.1)

#FOR HOLLINGS TYPE I
fit0_HT1_car_cap = optim(params_HT1_car_cap, rss_pi_HT1_car_cap, data=observed_df$green_data)
fit0_HT1_car_cap$par
#fit0$convergence
fit1_HT1_car_cap = optim(fit0_HT1$par, rss_pi_HT1_car_cap, data=observed_df$green_data)
fit1_HT1_car_cap$par
fit3_HT1_car_cap = optim(fit1_HT1_car_cap$par, rss_pi_HT1_car_cap, data=observed_df$green_data)
fit3_HT1_car_cap$par
```
```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)

pi_results_min_HT1_car_cap = ode(init_vals, times, pi_model_HT1_car_cap, fit0_HT1_car_cap$par)

pi_results_min_df_HT1_car_cap = data.frame(pi_results_min_HT1_car_cap)
#pi_results_min_df_HT1_car_cap

ggplot() + geom_line(aes(x=pi_results_min_df_HT1_car_cap$time, y=pi_results_min_df_HT1_car_cap$P, color='Optimized Parameters'))+geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```


```{r}
params_HT2_car_cap = c(beta=0.11, d = -0.2, p = 0.3 , B = -0.005, a = 2.6, c = 0.7, K=1000, r = 0.01, b=0.1)


#FOR HOLLINGS TYPE II
fit0_HT2_car_cap = optim(params_HT2_car_cap, rss_pi_HT2_car_cap, data=observed_df$green_data)
fit0_HT2_car_cap$par
#fit0$convergence
fit1_HT2_car_cap = optim(fit0_HT2_car_cap$par, rss_pi_HT2_car_cap, data=observed_df$green_data)
fit1_HT2_car_cap$par
fit3_HT2_car_cap = optim(fit1_HT2_car_cap$par, rss_pi_HT2_car_cap, data=observed_df$green_data)
fit3_HT2_car_cap$par
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.05)

pi_results_min_HT2_car_cap = ode(init_vals, times, pi_model_HT2_car_cap, fit3_HT2_car_cap$par)

pi_results_min_df_HT2_car_cap = data.frame(pi_results_min_HT2_car_cap)
#pi_results_min_df_HT2

ggplot() + geom_line(aes(x=pi_results_min_df_HT2_car_cap$time, y=pi_results_min_df_HT2_car_cap$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```

```{r}
params_HT3_car_cap = c(beta=3, d = -0.2, p = 0.3 , B = -0.005, a = 2.6, c = 0.7, K=1000, r = 0.01, b=0.1)

#FOR HOLLINGS TYPE III
fit0_HT3_car_cap = optim(params_HT3_car_cap, rss_pi_HT3_car_cap, data=observed_df$green_data)
fit0_HT3_car_cap$par
#fit0$convergence
fit1_HT3_car_cap  = optim(fit0_HT3_car_cap$par, rss_pi_HT3_car_cap, data=observed_df$green_data)
fit1_HT3_car_cap$par
fit3_HT3_car_cap = optim(fit1_HT3_car_cap$par, rss_pi_HT3_car_cap, data=observed_df$green_data)
fit3_HT3_car_cap$par
```

```{r}
times = observed_df$green_data_day #times of measurements from observed digitized data
  # Initial conditions
init_vals = c(P= observed_df$green_data_parasite[1], I = 0.1)

pi_results_min_HT3 = ode(init_vals, times, pi_model_HT3_car_cap, fit0_HT3_car_cap$par)

pi_results_min_df_HT3 = data.frame(pi_results_min_HT3)
pi_results_min_df_HT3

ggplot() + geom_line(aes(x=pi_results_min_df_HT3$time, y=pi_results_min_df_HT3$P, color='Optimized Parameters'))+
      geom_point(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data")) +
            geom_line(aes(x=observed_df$green_data_day, y=observed_df$green_data_parasite, color="Control - Experimental Obsv. Parasite Data"))+
  ylab("Population") + xlab("Time")
```








